---
title: "Prediction Assignment"
author: "AbdulAziz Ascandari"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

In recent years, the proliferation of wearable devices such as Jawbone Up, Nike FuelBand, and Fitbit has made it increasingly easy and affordable to collect extensive data on personal activities. These devices have become integral tools for the quantified self movement—a community of individuals who systematically track various aspects of their lives. The primary motivations for this practice include improving health, identifying behavioral patterns, and exploring technology. While there is a wealth of data available on the quantity of activities performed, there is a notable gap in quantifying the quality or effectiveness of these activities. 


## Aim

In this project, the objective is to utilize data collected from accelerometers placed on the belt, forearm, arm, and dumbbell of six participants. These participants were instructed to perform barbell lifts, both correctly and incorrectly, in five distinct ways. The analysis and modeling for this project will be conducted using the h2o library, a powerful tool for machine learning.The data for the analysis is taken from <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har>.

## Analysis

### Preprocessing

### Load Libraries

Here, we shall employ the h2o library for our prediction.

```{r}
library(h2o)
library(caret)
library(ggplot2)
library(dplyr)
library(tidyr)
```

### Load and clean data

```{r}
set.seed(1234)
training <- read.csv("pml-training.csv")

# Remove near-zero variance features
nzv <- nearZeroVar(training)
training_clean <- training[, -nzv]

# Remove columns with more than 95% missing values
na_cols <- sapply(training_clean, function(x) mean(is.na(x))) > 0.95
train_clean <- training_clean[, !na_cols]

# Remove irrelevant columns (e.g., identifiers)
train_clean <- train_clean[, -c(1:7)]  # Adjust the range as needed
```

### Data Splitting

```{r}
# Initialize H2O cluster
h2o.init()

# Convert data to H2O frame
train_h2o <- as.h2o(train_clean)

# Split into train, validation, and test sets (70%, 15%, 15%)
splits <- h2o.splitFrame(data = train_h2o, ratios = c(0.7, 0.15), seed = 1234)
train <- h2o.assign(splits[[1]], "train")
valid <- h2o.assign(splits[[2]], "valid")
test  <- h2o.assign(splits[[3]], "test")
```

#  Model Training

### Gradient Boosting


```{r}
# Convert the response column 'classe' to a factor (categorical variable)
train$classe <- as.factor(train$classe)
valid$classe <- as.factor(valid$classe)

y <- "classe"
x <- setdiff(names(train), y)

# Train GBM model
gbm_model <- h2o.gbm(
  x = x, y = y,
  training_frame = train,
  validation_frame = valid,
  ntrees = 500, max_depth = 6, learn_rate = 0.1,
  seed = 1234
)
```

### Random Forest

```{r}
rf_model <- h2o.randomForest(
  x = x, y = y,
  training_frame = train,
  validation_frame = valid,
  ntrees = 500, max_depth = 6,
  seed = 1234
)
```

### Naive Bayes
```{r}
nb_model <- h2o.naiveBayes(
  x = x, y = y,
  training_frame = train,
  validation_frame = valid,
  nfolds = 10,
  seed = 1234
)
```

### XGBoost

```{r}
xgb_model <- h2o.xgboost(
  x = x, y = y,
  training_frame = train,
  validation_frame = valid,
  seed = 1234
)
```

# Model Evaluation

### Compare R-squared for Each Model

```{r}
rsquare <- data.frame(
  Model = c("GBM", "RF", "NB", "XGBoost"),
  R2 = c(
    h2o.r2(gbm_model, valid = TRUE),
    h2o.r2(rf_model, valid = TRUE),
    h2o.r2(nb_model, valid = TRUE),
    h2o.r2(xgb_model, valid = TRUE)
  )
)

# Plot the R-squared values
ggplot(rsquare, aes(x = reorder(Model, -R2), y = R2)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Model Comparison (R-squared)", x = "Model", y = "R-squared")
```

From the R^2 plot the GBM model is selected. The confusion matrix is then printed.

### Confusion matrix of the selected Model (GBM)

```{r}
h2o.confusionMatrix(gbm_model, valid = TRUE)
```

### Variable Importance Plot (GBM)

```{r}
h2o.varimp_plot(gbm_model)
```

# Model Testing

## Load Test Data

```{r}
test_data <- read.csv("pml-testing.csv")
test_h2o <- as.h2o(test_data)
```

## Predict Using the GBM Model

```{r}
predictions <- h2o.predict(gbm_model, test_h2o)
predicted_classes <- as.data.frame(predictions$predict)
print(predicted_classes)
```

